DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2022-05-23 manuscript/manuscript.Rmd

DateLife's core software application consists of the R package `datelife`. Its latest stable version -- v`r packageVersion("datelife")`, is available from the CRAN repository [@datelifeR], and relies on functionalities from various biological R packages:
ape [@paradis2004],
bold [@Chamberlain2018],
geiger [@Harmon2008],
paleotree [@Bapst2012a],
phyloch [@Heibl2008],
phylocomr [@Ooms2018],
phytools [@Revell2012],
rotl [@Michonneau2016], and
taxize [@Chamberlain2013; @Chamberlain2018].
Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: providing an input, searching a chronogram database, and summarizing results from the search.

<!--_**Providing an input.--**_ -->

## Processing an input

@@

Alternative options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the secondary calibrations.
PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees.
treePL [@Smith2012], is a semi-parametric, rate-smoothing, penalized likelihood dating method
[@Sanderson2002].
The MrBayes  [@Huelsenbeck2001; @Ronquist2003] approach in datelife uses the secondary calibrations as priors on node ages.

@@

We illustrate the DateLife algorithm using a group within the Passeriform birds encompassing the family of true finches, Fringillidae, as case study. The first example analyses 6 bird species and shows all steps of the algorithm. The second example is an application analysing all 289 species in the family Fringillidae for which we have date informtaion.
<!--We used `r cite_r("r-references.bib")` for all our analyses.-->

@@

We randomly chose 6 bird species in the family Fringillidae, the true finches. The sample includes
two species of cardinals:
the black-thighed grosbeak -- *Pheucticus tibialis* and
the crimson-collared grosbeak -- *Rhodothraupis celaeno*;
three species of buntings:
the yellowhammer -- *Emberiza citrinella*,
the pine bunting --  *Emberiza leucocephalos* and
the yellow-throated bunting --  *Emberiza elegans*;
and one species of tanager, the vegetarian finch -- *Platyspiza crassirostris*.

Processing input names found that *Emberiza elegans* is synonym for *Schoeniclus elegans* in the default reference taxonomy [Open Tree of Life Taxonomy v3.3, June 1, 2021]. For a detailed discussion on the state of the synonym refer to Avibase [@lepage2004avibase; @lepage2014avibase; @avibase-emberiza]. 
DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies. Three studies matched five input names [@barker2015new; @Hedges2015; @Jetz2012], one study matched four input names [@hooper2017chromosomal] and two studies matched two input names [@barker2013going; @burns2014phylogenetics]. No studies matched all input names. Together, matching chronograms have 28 unique age data points. All nodes have age data.
DateLife used OpenTree's synthetic tree topology for these taxa and mapped age data to nodes in the tree.
As expected, more inclusive nodes (e.g., node "n1") have more age data than less inclusive nodes (e.g., node "n5").
The name resolution step allowed discovering five data points for node "n4" that would not have had any data otherwise due to name mismatch. 
Age summary statistics per node were calculated and used as secondary calibrations to date the tree topology using the BLADJ algorithm.
Age data for node "n2" was excluded as final calibration because it is older than age data of a more inclusive node (n1).

@@

To estimate ages for the taxa in the family Fringillidae, we ran a `datelife` query using the "get species from taxon" flag. 
This will get all recognised species names within the named taxon. 
The Fringillidae has species, according to the Open Tree of Life taxonomy. <!--BCO: Feel like we need a number here, though it's good that the clade hasn't gone extinct ;-) -->
??? We have date estimates for `r length(dquery$cleaned_names)` of those species.

??? There is an option to include all, right?
According to the OpenTree of life synthetic phylogeny, Fringillidae is not monophyletic, so this taxon-constrained approach returns a topology and ages for the species in that named group. <!--BCO: it's ambiguous: it gets a topology and age for the larger clade that includes all Fringillidae species? Or for the paraphyletic Fringillidae group?-->

The next step is to run a `datelife` search on this curated set of species taxon names that will find all chronograms that contain at least two species. The algorithm proceeds to prune the trees to keep matching species names on tips only, and transform the pruned trees to pairwise distance matrices.
There are `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
The final step is to summarize the available information using two alternative types of summary chronograms, median and SDM. As explained in the "Description" section, data from source chronograms is first summarised into a single patristic distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig. \ref{fig:fringillidages}). Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
<!-- ????????????????? Say some things about the results!
ANSWER:
(Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
<!-- and hierarchical clustering method which appears to push ages back -->

@@

<!--Data from source chronograms can be used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates-->
We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms. 
We used the individual tree topologies from each of the ??? TODO CHECK NUMBER 19 source chronograms from ?? studies as inputs, treating their node ages as unknown.
We estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.

We found that node ages from original study, and ages estimated using all other age data available are correlated  \textcolor{red}{R^2 value?} (Supplementary Fig. \ref{fig:cvXY}).

@@

Summarizing chronograms might also imply summarizing fundamentally distinct evolutionary hypotheses.
For example, two different researchers working on the same clade could both carefully select and argue for their choices of fossil calibrations.
Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses the same one to calibrate outside the clade, the resulting age estimates will often differ substantially, as the placement of calibrations as stem or crown group is proved to deeply affect estimated times of lineage divergence [@sauquet2013practical]. Trying to summarize the resulting chronograms into a single one using simple summary statistics can erase many types of relevant information from the source chronograms. Accordingly, the prevailing view is that we should favor lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history, as for example done by [@antonelli2017supersmart]. <!-- Will this be implemented in future `datelife` versions? Probably not. -->
There have been attempts to create fossil calibration databases <!--REF TO ADD: Ksepka, D.T., J.F. Parham, J.F. Allman, M.J. Benton, M.T. Carrano, K.A. Cranston, P.C. J. Donoghue, J.J. Head, E.J. Hermsen, R.B. Irmis, W.G. Joyce, M. Kohli, K.S. Lamm, D. Leehr, J.S.L. Patané, P.D. Polly, M.J. Phillips, N.A. Smith, N.D. Smith, M. van Tuinen, J.L. Ware, R.C.M. Warnock. 2015.The Fossil Calibration Database, A New Resource for Divergence Dating. Systematic Biology: doi:10.1093/sysbio/syv025.-->, though these still have room to grow. <!--BCO: Polite way to say 'no new additions in four years?'-->
However, the exercise of summarizing different chronograms has the potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypotheses. Choosing the elements of the chronograms that we are going to keep and the ones that we are going to discard is key, since we are potentially losing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram [@sauquet2021age]. <!-- Example -->
\textcolor{red}{??? it's not clear to me what this last part means! What might you be losing?}

>>>>

DateLife's core software application consists of the R package `datelife`. Its latest stable version -- v`r packageVersion("datelife")`, is available from the CRAN repository [@datelifeR], and relies on functionalities from various biological R packages:
ape [@paradis2004],
bold [@Chamberlain2018],
geiger [@Harmon2008],
paleotree [@Bapst2012a],
phyloch [@Heibl2008],
phylocomr [@Ooms2018],
phytools [@Revell2012],
rotl [@Michonneau2016], and
taxize [@Chamberlain2013; @Chamberlain2018].
Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: creating a search query, searching a database, and summarizing results from the search.

<!--_**Processing an input.--**_ -->

##  Creating a search query

@@

Alternative options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the secondary calibrations.
PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees.
treePL [@Smith2012], is a semi-parametric, rate-smoothing, penalized likelihood dating method
[@sanderson2002].
The MrBayes  [@Huelsenbeck2001; @Ronquist2003] approach in datelife uses the secondary calibrations as priors on node ages.

@@

We illustrate the DateLife algorithm using a family within the Passeriform birds encompassing the true finches, Fringillidae, as case study. The first example analyses 6 bird species and shows all steps of the algorithm. The second example is an application analysing 289 species in the family Fringillidae that are included in the NCBI taxonomy.
<!--We used `r cite_r("r-references.bib")` for all our analyses.-->

@@

_**Creating a search query.-**_
We chose 6 bird species in the family Fringillidae, known as the true finches. The sample includes
two species of cardinals:
the black-thighed grosbeak -- *Pheucticus tibialis* and
the crimson-collared grosbeak -- *Rhodothraupis celaeno*;
three species of buntings:
the yellowhammer -- *Emberiza citrinella*,
the pine bunting --  *Emberiza leucocephalos* and
the yellow-throated bunting --  *Emberiza elegans*;
and one species of tanager, the vegetarian finch -- *Platyspiza crassirostris*.

Processing input names found that *Emberiza elegans* is synonym for *Schoeniclus elegans* in the default reference taxonomy [Open Tree of Life Taxonomy v3.3, June 1, 2021]. For a detailed discussion on the state of the synonym refer to Avibase [@lepage2004avibase; @lepage2014avibase; @avibase-emberiza].
Discovering this synonym allowed assigning five age data points for the parent node of *Emberiza elegans*, shown as *Schoeniclus elegans* in figure \ref{fig:figure2-1}, which would not have had any data otherwise.
<!--\textcolor{red}{I don't understand what this sentence means}
it is related to the synonym found by tnrs processing, if we had used the synonym insted of the one accepted in the database, we would not have discovered that information
It moved it here, it makes more sense here.
-->

_**Searching the database.-**_
DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies. Three studies matched five input names [@barker2015new; @Hedges2015; @Jetz2012], one study matched four input names [@hooper2017chromosomal] and two studies matched two input names [@barker2013going; @burns2014phylogenetics]. No studies matched all input names. Together, matching chronograms have 28 unique age data points. All nodes have age data.

_**Summarizing search results.-**_
DateLife used OpenTree's synthetic tree topology for these taxa and mapped age data to nodes in the tree.
As expected, more inclusive nodes (e.g., node "n1") have more age data than less inclusive nodes (e.g., node "n5").
The name resolution step allowed discovering five data points for node "n4" that would not have had any data otherwise due to name mismatch.
Age summary statistics per node were calculated and used as secondary calibrations to date the tree topology using the BLADJ algorithm.
Age data for node "n2" was excluded as final calibration because it is older than age data of the more inclusive node "n1".

@@

_**Creating a query.-**_
To estimate ages for species in the family Fringillidae, we ran a `datelife` query using the "get species from taxon" flag,
which gets all recognized species names within a named group from a taxonomy of choice (options are Open Tree of Life, NCBI, GBIF, or IRMNG).
Following the NCBI taxonomy, the Fringillidae has `r length(dquery$cleaned_names)` species.
<!--BCO: Feel like we need a number here, though it's good that the clade hasn't gone extinct ;-) -->

_**Searching the database.-**_
The next step is to use this curated set of species taxon names to identify all chronograms that contain at least two Fringillidae species. Once identified, the algorithm proceeds to prune matching chronograms to keep Fringillidae species names on tips only, and transform these pruned chronograms to pairwise distance matrices.
The `datelife` search revealed `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].

_**Summarizing search results.-**_
The final step is to summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM. As explained in the "Description" section, a tree topology to summarize age data upon is required.
By default, `datelife` uses the Open Tree of Life synthetic phylogeny as summarizing topology. According to this phylogeny, Fringillidae is not a monophyletic family
[@burns2014phylogenetics;
@burleigh2015building;
@Jetz2012;
@lerner2011multilocus;
@lovette2010comprehensive;
@pulgarin2013multilocus;
@chaves2013biogeography;
@lamichhaney2015evolution;
@parchman2007coevolution;
@selvatti2015paleogene;
@moyle2016tectonic;
@odeen2011evolution;
@claramunt2015new;
@beresford2005african;
@kimball2019phylogenomic;
@barker2015new;
@klicka2014comprehensive;
@barker2013going;
@treplin2008molecular;
@price2014niche;
@hackett2008phylogenomic;
@barker2014mitogenomic;
@gibb2015new;
@barker2004phylogeny;
@zuccon2012phylogenetic;
@oliveros2019earth;
@johansson2008phylogenetic;
@alstrom2014discovery;
@powell2014comprehensive;
@bryson2014diversification;
@packert2012horizontal;
@tietze2013complete].

Hence, DateLife's taxon-constrained approach returns a topology and ages for the species in a named group.

<!--BCO: it's ambiguous: it gets a topology and age for the larger clade that includes all Fringillidae species? Or for the paraphyletic Fringillidae group?-->

Age data from source chronograms is summarised into a single patristic distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig. \ref{fig:fringillidages}).

Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
<!-- ????????????????? Say some things about the results!
ANSWER:
(Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
<!-- and hierarchical clustering method which appears to push ages back -->

@@

<!--Data from source chronograms can be used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates-->
We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms.
We used the individual tree topologies from each of the 19 source chronograms from 13 studies as inputs, treating their node ages as unknown.
We then estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.

We found that node ages from original study, and ages estimated using all other age data available are correlated (Supplementary Fig. \ref{fig:cvXY}).

@@

Summarizing chronograms might also imply summarizing fundamentally distinct evolutionary hypotheses.
For example, two different researchers working on the same clade could both carefully select and argue for their choices of fossil calibrations.
Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses the same one to calibrate outside the clade, the resulting age estimates will often differ substantially, as the placement of calibrations as stem or crown group is proved to deeply affect estimated times of lineage divergence [@sauquet2013practical]. Trying to summarize the resulting chronograms into a single one using simple summary statistics can erase many types of relevant information from the source chronograms. Accordingly, the prevailing view is that we should favor lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history, as for example done by [@antonelli2017supersmart]. <!-- Will this be implemented in future `datelife` versions? Probably not. -->
There have been attempts to create fossil calibration databases <!--REF TO ADD: Ksepka, D.T., J.F. Parham, J.F. Allman, M.J. Benton, M.T. Carrano, K.A. Cranston, P.C. J. Donoghue, J.J. Head, E.J. Hermsen, R.B. Irmis, W.G. Joyce, M. Kohli, K.S. Lamm, D. Leehr, J.S.L. Patané, P.D. Polly, M.J. Phillips, N.A. Smith, N.D. Smith, M. van Tuinen, J.L. Ware, R.C.M. Warnock. 2015.The Fossil Calibration Database, A New Resource for Divergence Dating. Systematic Biology: doi:10.1093/sysbio/syv025.-->, though these still have room to grow. <!--BCO: Polite way to say 'no new additions in four years?'-->
However, the exercise of summarizing different chronograms has the potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypotheses. Choosing the elements of the chronograms that we are going to keep and the ones that we are going to discard is key, since we are potentially losing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram [@sauquet2021age].
<!-- Example
\textcolor{red}{EJM: ??? it's not clear to me what this last part means! What might you be losing?}
-->
